---
title: "Proj2red"
author: "Walter Coleman"
date: "8/10/2021"
output: pdf_document
---
```{r}
setwd('/Users/colemanwalterj/Documents/OneDrive/UVA MS Data Science/Linear Models for DS')
library(tidyverse)
library(ggplot2)
library(leaps)
red<-read.csv('wineQualityReds.csv',header=TRUE)
red<-select(red, -(X))

##set the random number generator so same results can ##be reproduced
set.seed(1)
##choose the observations to be in the training set.
##I am splitting the data set into halves 
samplered<-sample.int(nrow(red), floor(.80*nrow(red)), replace = F) 
trainred<-red[samplered, ] ##training data
testred<-red[-samplered, ] ##test data

head(trainred)
```
```{r}
ggplot(trainred, aes(x=residual.sugar, y=alcohol,color=quality))+
  geom_point()+
  labs(x="Residual Sugar", y="Alcohol", title="Alcohol against Residual Sugar with Quality")
```
```{r}
ggplot(trainred, aes(x=citric.acid, y=pH,color=alcohol))+
  geom_point()+
  labs(x="Citric Acid", y="pH", title="pH against Citric Acid with Quality")
```

```{r}
library(gridExtra)
bp1<-ggplot(trainred, aes(x=fixed.acidity, y=quality))+
  geom_boxplot()+
  labs(x="fixed.acidity", y="quality",title="quality against fixed.acidity")
bp2<-ggplot(trainred, aes(x=volatile.acidity, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="volatile.acidity", y="quality",title="quality against volatile.acidity")
bp3<-ggplot(trainred, aes(x=citric.acid, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="citric.acid", y="quality",title="quality against citric.acid")
bp4<-ggplot(trainred, aes(x=residual.sugar, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="residual.sugar", y="quality",title="quality against residual.sugar")
bp5<-ggplot(trainred, aes(x=chlorides, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="chlorides", y="quality",title="quality against chlorides")
bp6<-ggplot(trainred, aes(x=free.sulfur.dioxide, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="free.sulfur.dioxide", y="quality",title="quality against free.sulfur.dioxide")

##produce the 4 boxplots in a 2 by 2 matrix
grid.arrange(bp1, bp2, bp3, bp4, bp5, bp6, ncol = 2, nrow = 3)
```
```{r}
bp7<-ggplot(trainred, aes(x=total.sulfur.dioxide, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="total.sulfur.dioxide", y="quality",title="quality against total.sulfur.dioxide")
bp8<-ggplot(trainred, aes(x=density, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="density", y="quality",title="quality against density")
bp9<-ggplot(trainred, aes(x=pH, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="pH", y="quality",title="quality against pH")
bp10<-ggplot(trainred, aes(x=sulphates, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="sulphates", y="quality",title="quality against sulphates")
bp11<-ggplot(trainred, aes(x=alcohol, y=quality))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="alcohol", y="quality",title="quality against alcohol")

grid.arrange(bp7, bp8, bp9, bp10, bp11, ncol = 2, nrow = 3)
```

```{r}
pairs(trainred, lower.panel = NULL)
```

```{r}
result<-lm(quality ~ ., data = trainred)
summary(result)
```

We test now if at least one of the coefficients is not zero:

H_0: beta1=beta2=...beta11=0
H_a: at least one betaj is not zero for j in 1,...,11

With F-statistic 81.35 and a low p-value, we reject the null hypothesis and conclude that at least one coefficient is not zero.

We believe that given the insignificant results of fixed.acidity, citric.acid, residual.sugar, density, pH we may be better off dropping them from our model. Thus, we compare with a reduced linear regression model and run ANOVA to test if we can drop.
```{r}
reduced<-lm(quality ~ volatile.acidity+chlorides+free.sulfur.dioxide+total.sulfur.dioxide+sulphates+alcohol, data=trainred)

anova(reduced, result)
```

H_0: beta1 = beta4 = beta5 = beta9 = beta10 = 0
H_a: at least one betaj is not zero for j in 1,4,5,9,10

With F statistic 2.74 and a low p-value, we reject the null hypothesis and do not drop the predictors neglected in our reduced model.

Let's also check for multicollinearity. Although our p-values are mostly significant (as is our p-value), we still want to see if any variables are highly correlated:
```{r}
round(cor(trainred),3)
```
```{r}
vif(result)
```
None of these predictors have Variance Inflation Factors >10, so we do not see multicollinearity. 

Adjusted R-squared:
```{r}
allreg <- regsubsets(quality ~., data=trainred, nbest=1)
summary(allreg)
```
```{r}
which.max(summary(allreg)$adjr2)
```
```{r}
coef(allreg, which.max(summary(allreg)$adjr2))
```


We will also run a forward selection to see the best model based on AIC:
```{r}
##intercept only model
regnull <- lm(quality~1, data=trainred) 
##model with all predictors 
regfull <- lm(quality~., data=trainred)

step(regnull, scope=list(lower=regnull, upper=regfull), direction="forward")
```

And backwards elimination:
```{r}
step(regfull, scope=list(lower=regnull, upper=regfull), direction="backward")
```

Stepwise regression for direction set to both:
```{r}
step(regfull, scope=list(lower=regnull, upper=regfull), direction="both")
```

Adjusted R^2:
quality ~ volatile.acidity + chlorides + free.sulfur.dioxide + total.sulfur.dioxide + pH + sulphates + alcohol 

Forward Selection:
quality ~ volatile.acidity + chlorides + free.sulfur.dioxide + total.sulfur.dioxide + pH + sulphates + alcohol

Backward Elimination:
quality ~ volatile.acidity + chlorides + free.sulfur.dioxide + total.sulfur.dioxide + pH + sulphates + alcohol

Stepwise Regression:
quality ~ volatile.acidity + chlorides + free.sulfur.dioxide + total.sulfur.dioxide + pH + sulphates + alcohol

All stepwise regression direction settings yielded the same model based on AIC, which was also the best model for Adjusted R^2:

quality = 4.089579 + 0.299231(alcohol) - 0.962398(volatile.acidity) + 0.881832(sulphates) - 0.003293(total.sulfur.dioxide) - 1.585888(chlorides) - 0.436359(pH) +  0.005371(free.sulfur.dioxide)

Attempting partial F test again after forward selection AIC-based recommendation, 
```{r}
reduced2 <- lm(quality ~ alcohol + volatile.acidity + sulphates + 
    total.sulfur.dioxide + chlorides + pH + free.sulfur.dioxide, 
    data = trainred)

anova(reduced2, result)
```

With F-statistic 0.7663 and a high p-value, we fail to reject the null hypothesis and drop the neglected predictors in our second reduced model.


We then check our new reduced model to see if any individual predictor can be dropped:
```{r}
summary(reduced2)
```
Given a significant F statistic and significant predictors, we will reduce our model no further. 

We will now use our reduced model to make prediction on the test data and calculate mean-squared error:
```{r}
preds<-predict(reduced2,newdata=testred)

## Mean squared error
sum((testred$quality-preds)^2)/nrow(testred)
```
